Conservator's Corner

### 2020-03-08 pnsac/newsletter/article_EditorsNotes.tex
PNSAC newsletter--Mar.2020.

Putting out this edition of the Chronicle has been an example of great
co-operation from members and non-members alike. Thanks are owed to a
number of people. Our member Gary Whitten was instrumental in getting
the necessary permission from Air Force Magazine to publish the
"Diversion to Alert" article.  Allan Bowes, another member, suggested
that we should try to get the "Civil Merlins" article into an edition
of the Chronicle.  Following his suggestion we contacted Ben Dunnell,
the editor of Aeroplane Magazine. Through Ben's offices we were able to
secure permission to reprint this article.  The "Civil Merlins" was
part of a comprehensive 17 page Aeroplane Database article on the
Merlin. If any members wish to read the entire article it was published
in the July 2017 issue.

In addition thanks go out to the other contributors, namely Richard
Lodge, Garry Dupont,  Neil Raynor and Bruce Gemmill. Bruce took the
lead in putting together the Progress Report.  Finally, I would like to
thank Drew Hodge our stalwart publisher of this edition. We are always
on the lookout for content for the Chronicle so please do not hesitate
to bring forward any ideas that you might have. Enjoy.

>>>>

Putting out this edition of the Chronicle has been an example of great
co-operation from members and non-members alike. Thanks are owed to a
number of people. Our member Gary Whitten was instrumental in getting
the necessary permission from Air Force Magazine to publish the
"Diversion to Alert" article.

%Allan Bowes, another member, suggested
%that we should try to get the "Civil Merlins" article into an edition
%of the Chronicle.  Following his suggestion we contacted Ben Dunnell,
%the editor of Aeroplane Magazine. Through Ben's offices we were able to
%secure permission to reprint this article.  The "Civil Merlins" was
%part of a comprehensive 17 page Aeroplane Database article on the
%Merlin. If any members wish to read the entire article it was published
%in the July 2017 issue.

In addition thanks go out to the other contributors, namely Richard
Lodge, Garry Dupont,  Neil Raynor, and R\'{e}jean Demers. A special thanks
to Bruce Gemmill who took the lead in putting together the progress report,
which can be found in the Conservator's Corner section.

Finally, I would like to thank Drew Hodge our stalwart publisher of this
edition. We do appreciate receiving feedback from our readers including
comments on the articles or suggestions for new articles. We are always on the
lookout for content for the Chronicle so please do not hesitate to bring
forward any ideas that you might have. As noted in the Conservator's Corner
article the restoration of engine number 4 has been completed. Our next
edition will focus on the engine restoration work so if you have any material
or comments specific to that aspect of the restoration please let us know.
Enjoy.

### 2020-03-08 pnsac/newsletter/article_Feature1.tex
PNSAC newsletter--Mar.2020.

Immediately after New Year 1960, while based at RCAF Stn Trenton, and
flying North Stars for 426(T)Sqn., I was assigned to be the First
Officer on a flight to pick up an entertainment troupe of 30 Passengers
from Montreal. This was planned as an 8 day trip providing
entertainment to personnel working along the Quebec, Ontario and
Manitoba sectors of the Mid Canada Detection Line.

@@

Our 426Sqn crew consisted of Captain – F/O H.(Hank) Zbesheski, First
Officer – F/O M.G.(Mac or Mo) Morrison, Navigator – F/O Roger Fridel,
Radio Officer – F/O Peter Dragovich, Flt Engineer – Sgt Miller, Flt
Engineer – Sgt Sopaz, Trans Tech – LAC Hall.

@@

@@

@@

@@

@@

>>>>

Immediately after New Year 1960, while based at RCAF Stn Trenton, and
flying North Stars for 426(T) Sqn., I was assigned to be the First
Officer on a flight to pick up an entertainment troupe of 30 Passengers
from Montreal. This was planned as an 8 day trip providing
entertainment to personnel working along the Quebec, Ontario and
Manitoba sectors of the Mid Canada Detection Line.

\begin{figure}[httb]
   \vspace{2em}
   \centering
   \includegraphics [scale=0.5]{alet-PC-155.png}
   \caption*{\small \em The venerable Canadair North Star provided yeoman service to the RCAF---RCAF Photo PC-155}
   \label{fig:wall-two}
\end{figure}

@@

Our 426Sqn crew consisted of Captain---F/O H.(Hank) Zbesheski, First
Officer---F/O M.G.(Mac or Mo) Morrison, Navigator---F/O Roger Fridel,
Radio Officer---F/O Peter Dragovich, Flt Engineer---Sgt Miller, Flt
Engineer---Sgt Sopaz, Trans Tech---LAC Hall.

@@

\begin{figure}[httb]
   \vspace{2em}
   \centering
   \includegraphics [scale=0.5]{alert-PCP-83-191_cropped.png}
   \caption*{\small \em North Stars were routinely used in harsh Arctic conditions---RCAF Photo PCP-83-191}
   \label{fig:wall-two}
\end{figure}

@@

\begin{figure}[httb]
   \vspace{2em}
   \centering
   \includegraphics [scale=0.5]{alert-PL-54044.png}
   \caption*{\small \em One of the North Stars from the Morrison story---aircraft \#17512---RCAF Photo PL-54044}
   \label{fig:wall-two}
\end{figure}

@@

%\begin{figure}[httb]
%   \vspace{2em}
%   \centering
%   \includegraphics [scale=0.5]{alert-PL-54068.png}
%   \caption*{\small \em The beauty of Canada's Arctic is framed by the nose of a North Star---RCAF Photo PL-54068}
%   \label{fig:wall-two}
%\end{figure}

@@

\end{multicols}

\begin{figure}[httb]
   \vspace{2em}
   \centering
   \includegraphics {alert-PL-54068.png}
   \caption*{\small \em The beauty of Canada's Arctic is framed by the nose of a North Star---RCAF Photo PL-54068}
   \label{fig:wall-two}
\end{figure}

@@

\begin{multicols}{2}

### 2020-03-08 pnsac/newsletter/article_Feature2.tex
PNSAC newsletter--Mar.2020.

The earliest Merlins to operate in a civil mode were the Merlin T24
series, developed in 1944. These were single-stage twin-speed units
similar to the Merlin 24s fitted to the RAF's Lancasters, but were
modified to improve service life, and were fitted to Transport
Command's Avro Yorks. Long-range transport operation entailed running
at relatively low cruise powers for long periods — under these
conditions the lower cylinder head temperatures caused deposits of lead
oxide from the fuel, resulting in excessive spark plug fouling. To
counter this, the Merlin T24/4 incorporated a charge heater to increase
the inlet temperature. Post-war, the Merlin 500 series essentially
comprised civil and export versions of the T24, incorporating its
modifications, though only the Merlin 501 included a charge heater.

@@

>>>>

The earliest Merlins to operate in a civil mode were the Merlin T24
series, developed in 1944. These were single-stage twin-speed units
similar to the Merlin 24s fitted to the RAF's Lancasters, but were
modified to improve service life, and were fitted to Transport
Command's Avro Yorks. Long-range transport operation entailed running
at relatively low cruise powers for long periods---under these
conditions the lower cylinder head temperatures caused deposits of lead
oxide from the fuel, resulting in excessive spark plug fouling. To
counter this, the Merlin T24/4 incorporated a charge heater to increase
the inlet temperature. Post-war, the Merlin 500 series essentially
comprised civil and export versions of the T24, incorporating its
modifications, though only the Merlin 501 included a charge heater.

\begin{figure}[httb]
   \vspace{2em}
   \centering
   \includegraphics [scale=0.5]{merlin-1-feb-2020.jpg}
   \caption*{\small \em Rolls-Royce Merlin Engine}
   \label{fig:wall-two}
\end{figure}

@@

\begin{figure}[httb]
   \vspace{2em}
   \centering
   \includegraphics[scale=0.5]{nstar-tca-feb-2020.jpg}
   \caption*{\small \em TCA Canadair DC-4M North Star}
   \label{fig:wall-two}
\end{figure}

### 2020-03-08 pnsac/newsletter/article_Feature3.tex
PNSAC newsletter--Mar.2020.

Seven years ago a group of people came together within the Multifaith
Housing Initiative (MHI) to do something about that stat. MHI is a
federally registered charity based in Ottawa which provides affordable
housing: currently it has about 140 units with 400 residents scattered
across the city. What started as a desire to provide affordable housing
has morphed into an ethos to build communities – Veterans' House is the
latest iteration of that approach.

@@

Together with our partners – many of them with direct connections to
the military and Veterans – Veterans' House: the Andy Carswell Building
will provide a "military-family" structure to residents which they will
recognize and which research has shown many homeless Veterans miss and
which contributes to their slide into homelessness. A fully list of our
partners, progress to date and lots of other information is on the MHI
Veterans' House website at
{\normalfont\color{blue}\texttt{\url{https://www.multifaithhousing.ca/veterans-house}}}

>>>>

Seven years ago a group of people came together within the Multifaith
Housing Initiative (MHI) to do something about that stat. MHI is a
federally registered charity based in Ottawa which provides affordable
housing: currently it has about 140 units with 400 residents scattered
across the city. What started as a desire to provide affordable housing
has morphed into an ethos to build communities---Veterans' House is the
latest iteration of that approach.

@@

Together with our partners---many of them with direct connections to
the military and Veterans---Veterans' House: the Andy Carswell Building
will provide a "military-family" structure to residents which they will
recognize and which research has shown many homeless Veterans miss and
which contributes to their slide into homelessness. A fully list of our
partners, progress to date and lots of other information is on the MHI
Veterans' House website at
{\normalfont\color{blue}\texttt{\url{https://www.multifaithhousing.ca/veterans-house}}}

### 2020-03-08 pnsac/newsletter/article_PresReport.tex
PNSAC newsletter--Mar.2020.

Since the last issue of the NStar Chronicle we have taken part in the
70th Anniversary celebrations of the first non-stop transcontinental
flight across Canada which was flown by a North Star. We partnered with
the Museum for this event and were very pleased to welcome
representatives from the RCAF who joined us for the day. After the
formal part of the celebration, when impressive plaques of the aircraft
made by our member Chris McGuffin, had been presented to the RCAF and
the Museum Director General, Chris Kitzan, everybody moved to the
Reserve Hangar and the aircraft was opened to our association members,
guests and in the afternoon to 90 members of the public who had
pre-booked a tour. R\'{e}jean(R\'{e}j)Demers, the Museum Special Project
Manager, guided the tours assisted by several of our members. Towards
the end of the tour session someone in the main Museum activated the
fire alarm. The Museum was promptly evacuated, and the public was
recommended to go to the Reserve  Hangar until the fire issue had been
dealt with (it turned out to be a false alarm). Quite unexpectedly R\'{e}j
and the guides found themselves surrounded by many Museum visitors who
had not signed up for the North Star tour. Unperturbed by this event,
Réj and the guides welcomed the unscheduled visitors and gave them an
impromptu tour of the aircraft. Both our treasurer and our membership
secretary were pleased with the days' result!

@@

This year we are planning two major display events. The first will be
on the Doors Open Ottawa weekend (June 6 and 7). This is a great time
for our members to meet the public, many of whom have an interest in
aviation. Our second main event, on Canada Day, is enjoyable for all
the members who sign up to be part of the team meeting the public.
There are always large crowds, many of them family groups, If any of
you reading this article would be interested in joining our members for
either of these events please contact me (my co-ordinates are on the
back page of this issue of the Chronicle)

Shortly before Christmas a major milestone, in the  restoration of the
North Star, was reached. Work on Merlin engine \#4 was completed and it
was dragged on its cradle by the Museum forklift truck to the
Restoration hangar for later mounting on the aircraft to join the other
three restored engines. Chris McGuffin put some video clips onto our
Facebook page of the engine being dragged through the Museum parking
lot and into the hangar. Watch for more of our Facebook messages since
Chris frequently posts new items.

>>>>

The notes below were written in the middle of February and since that
time a sad event has taken place in that our old friend and Life Member
of PNSAC, Ted Devey, has passed away. We have only a few Life Members.
Such memberships are granted only in circumstances where a member has
made a special and exceptional contribution to the North Star
restoration project. Ted Devey was one such member. At 94 years old,
Ted was our oldest member and the oldest Museum restoration volunteer.
We will miss Ted, his continuous supply of amusing anecdotes and of
course his dedication to the North Star restoration project. Ted was
the subject of "Our Members" section of the January 2019 edition of the
Chronicle which is available on the Association's website.

Since the last issue of the NStar Chronicle we have taken part in the
70th Anniversary celebrations of the first non-stop transcontinental
flight across Canada which was flown by a North Star. Each year work is
performed on the aircraft which cannot be done outside the Reserve Hangar
during the winter. The aircraft is also prepared for display to enable the
public to view the progress made in the last year. After the
formal part of the celebration, when impressive plaques of the aircraft
made by our member Chris McGuffin, had been presented to the RCAF and
the Museum Director General, Chris Kitzan, everybody moved to the
Reserve Hangar and the aircraft was opened to our association members,
guests and in the afternoon to 90 members of the public who had
pre-booked a tour. R\'{e}jean (R\'{e}j) Demers, the Museum Special Project
Manager, guided the tours assisted by several of our members. Towards
the end of the tour session someone in the main Museum activated the
fire alarm. The Museum was promptly evacuated, and the public was
recommended to go to the Reserve  Hangar until the fire issue had been
dealt with (it turned out to be a false alarm). Quite unexpectedly R\'{e}j
and the guides found themselves surrounded by many Museum visitors who
had not signed up for the North Star tour. Unperturbed by this event,
R\'{e}j and the guides welcomed the unscheduled visitors and gave them an
impromptu tour of the aircraft. Both our treasurer and our membership
secretary were pleased with the days' result!

% We partnered with
%the Museum for this event and were very pleased to welcome
%representatives from the RCAF who joined us for the day.

@@

This year the Museum is planning two major display events. The first will be
on the Doors Open Ottawa weekend (June 6 and 7). This is a great time
for our members to meet the public, many of whom have an interest in
aviation. Our second main event, on Canada Day, is enjoyable for all
the members who sign up to be part of the team meeting the public.
There are always large crowds, many of them family groups, If any of
you reading this article would be interested in joining our members for
either of these events please contact me (my co-ordinates are on the
back page of this issue of the Chronicle)

Shortly before Christmas a major milestone, in the  restoration of the
North Star, was reached. Work on Merlin engine \#4 was completed and it
was dragged on its cradle by the Museum forklift truck to the
Reserve hangar for later mounting on the aircraft to join the other
three restored engines. Chris McGuffin put some video clips onto our
Facebook page of the engine being dragged through the Museum parking
lot and into the hangar. Watch for more of our Facebook messages since
Chris frequently posts new items.

### 2020-03-08 pnsac/newsletter/article_ProgReport.tex
PNSAC newsletter--Mar.2020.

Most notably were efforts by Charles Baril, who spent many months
inside the nacelle; cleaning, inspecting and masking off for painting.
The work took so long that special heating ducts and insulation needed
to be positioned around the nacelle to keep it warm enough for Réjean
Demers and Bruce Gemmill to apply a coat of primer and silver paint.
(insert photo of North Star with orange tarp). Soon after the paint was
applied, the North Star was moved inside for the winter. This occurred
November 8th, just 3 days before our first snow storm of the winter.

@@

The Carpentry Crew, led by Chris McGuffin measured, cut and trimmed new
plywood panels for the interior walls of the main cabin.  These were
painted green to match the original colour scheme, then set aside until
for later installation.  Next came the floor panels.  Richard Houle and
John Makadi assessed these, according to the amount of damage caused by
wear and tear, as well as leaks around windows and doors.  Some panels
were salvaged with a few minor repairs, then painted. These are now
ready for installation.  Several others were so badly damaged that
completely new panels needed to be made. (Photo of wall and floor
panel)

@@

Many of the volunteers got to work on the large collection of
accessories that had been removed from nacelle \#4.  This included
pumps, plumbing, fluid and air lines, control cables, bell cranks and
pulleys, electrical harnesses, valves and filter canisters. Each needed
assessment, cleaning, repair, polishing or painting as well as
documentation and photographs of the work undertaken. (photo of fuel
filter, junction box).

Phil Chrysler and Brian Cole went to work on the many fairings and
access panels that also required cleaning and polishing to be ready to
be placed back on the aircraft. A special mention here for Ted Devey,
who, before retiring from the shop had his hand on a few cowl panels
and nacelle components.

Several special projects were also undertaken during the year.  John
Makadi removed the nose radome to determine what radar equipment had
been removed, in case we are able to find a source for missing
equipment.  We removed the upper rotating navigation beacon to fix a
persistent leak, this necessitated a twin tower scaffold assembly to
safely reach the top of the fuselage. Neil Raynor and Will Assad
overhauled a propeller stand the Museum had not used for many years. 
It got a fresh coat of paint and new wheels and tires, and is now
holding two North Star propellers, until they can be installed on the
appropriate engine. (photo of prop stand, two propellers). Michael Hope
assisted in removal of Prop. \#1 and assembly of soon-to-be-installed
Prop. \#4. Phil Chrysler and Brian Cole fabricated cables and special
fittings to secure the landing gear so the aircraft could be safely
displayed. Other duties  such as; an inventory of replaced / restored
and spare components were carried out.  All of these were packed and
stored in a new location of the Reserve Hangar.

@@

Currently, work is on-going to complete the rear (main) cargo door and
quilted liner installation. Nacelle \#4 installation of restored
components is advancing in small steps, due to the complexity of
interrelated systems and locations. Scaffolding is in place, over the
forward fuselage area above the cockpit to assess air-frame
requirements and treatment methods to be used on exterior hull
surfaces. Restoration records are also being reviewed by Conservation
Special Project Manager Réjean Demers. The Work Order / Task Card
format is lending well to Museum information uploading tools. Efforts
are underway to recruit two to four new shop volunteers, and R\'{e}j
invites any new applicants referred by PNSAC to apply.

>>>>

Most notably were efforts by Charles Baril, who spent many months
inside the nacelle; cleaning, inspecting and masking off for painting.
The work took so long that special heating ducts and insulation needed
to be positioned around the nacelle to keep it warm enough for R\'{e}jean
Demers and Bruce Gemmill to apply a coat of primer and silver paint.

\begin{figure}[httb]
   \vspace{2em}
   \centering
   \includegraphics [scale=0.5]{nacelle-4-paining-outside-feb-2020.png}
   \caption*{\small \em Painting nacelle \#4 outside}
   \label{fig:wall-two}
\end{figure}

Soon after the paint was
applied, the North Star was moved inside for the winter. This occurred
November 8th, just 3 days before our first snow storm of the winter.

@@

\begin{figure}[httb]
   \vspace{2em}
   \centering
   \includegraphics [scale=0.5]{New-Floor-Wall-Panels-feb-2020.png}
   \caption*{\small \em New floor and wall panels}
   \label{fig:wall-two}
\end{figure}

The Carpentry Crew, led by Chris McGuffin measured, cut and trimmed new
plywood panels for the interior walls of the main cabin.  These were
painted green to match the original colour scheme, then set aside until
for later installation.  Next came the floor panels.  Richard Houle and
John Makadi assessed these, according to the amount of damage caused by
wear and tear, as well as leaks around windows and doors.  Some panels
were salvaged with a few minor repairs, then painted. These are now
ready for installation.  Several others were so badly damaged that
completely new panels needed to be made.

@@

Many of the volunteers got to work on the large collection of
accessories that had been removed from nacelle \#4.  This included
pumps, plumbing, fluid and air lines, control cables, bell cranks and
pulleys, electrical harnesses, valves and filter canisters. Each needed
assessment, cleaning, repair, polishing or painting as well as
documentation and photographs of the work undertaken. Phil Chrysler and 
Brian Cole went to work on the many fairings and
access panels that also required cleaning and polishing to be ready to
be placed back on the aircraft. A special mention should be made for the late
Ted Devey who had had his hand on a few cowl panels and nacelle components.

Several special projects were also undertaken during the year.  John
Makadi removed the nose radome to determine what radar equipment had
been removed, in case we are able to find a source for missing
equipment.  We removed the upper rotating navigation beacon to fix a
persistent leak, this necessitated a twin tower scaffold assembly to
safely reach the top of the fuselage. Neil Raynor and Will Assad
overhauled a propeller stand the Museum had not used for many years. 
It got a fresh coat of paint and new wheels and tires, and is now
holding two North Star propellers, until they can be installed on the
appropriate engine.

\begin{figure}[httb]
   \vspace{2em}
   \centering
   \includegraphics [scale=0.5]{prop-stand.png}
   \caption*{\small \em Prop stand}
   \label{fig:wall-two}
\end{figure}

Michael Hope
assisted in removal of Prop. \#1 and assembly of soon-to-be-installed
Prop. \#4. Phil Chrysler and Brian Cole fabricated cables and special
fittings to secure the landing gear so the aircraft could be safely
displayed. Other duties  such as; an inventory of replaced / restored
and spare components were carried out.  All of these were packed and
stored in a new location of the Reserve Hangar.

\begin{figure}[httb]
   \vspace{2em}
   \centering
   \includegraphics [scale=0.5]{eng-4-pre-move-feb-2020.png}
   \caption*{\small \em Engine \#4---before it was moved to the Reserve Hangar}
   \label{fig:wall-two}
\end{figure}

@@

Currently, work is on-going to complete the rear (main) cargo door and
quilted liner installation. Nacelle \#4 installation of restored
components is advancing in small steps, due to the complexity of
interrelated systems and locations. Scaffolding is in place, over the
forward fuselage area above the cockpit to assess air-frame
requirements and treatment methods to be used on exterior hull
surfaces. Restoration records are also being reviewed by Conservation
Special Project Manager R\'{e}jean Demers. The Work Order / Task Card
format is lending well to Museum information uploading tools. Efforts
are underway to recruit two to four new shop volunteers, and R\'{e}j
invites any new applicants referred by PNSAC to apply.
